---
title: "GDP and lifexp"
author: "Zichun Ye"
date: "2014/09/26"
output: html_document
---
In this part, we will continue to explore the data set [Gapminder excerpt](http://www.stat.ubc.ca/~jenny/notOcto/STAT545A/examples/gapminder/data/gapminderDataFiveYear.txt). 

# Preparation

Before preceeding to the exciting parts, we need some preparation like loading the data and library.

```{r}
# load the data
gdURL <-"http://tiny.cc/gapminder"
gDat <- read.delim(file = gdURL)

# load the library
library(dplyr)
library(ggplot2)
library(ggthemes)
library(knitr)
library(reshape2)

# change data.frame to tbl_df 
gtbl <- tbl_df(gDat)
glimpse(gtbl)
```

# Our exploration
* TASK NO.1: Get the maximum and minimum of GDP per capita for all continents.
```{r}
# Get the maximum and minimum of GDP per capita for all continents.
gdp_int <- gtbl %>%
  group_by(continent) %>%
  summarize(min_gdpPercap = min(gdpPercap), max_gdpPercap = max(gdpPercap))

# resharp the data for plot
gdp_int.r = melt(gdp_int)
```

```{r two-column, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(gdp_int)
cat("</td>")
cat("<td>")
ggplot(gdp_int.r, aes(continent,value,fill=variable))+
  geom_bar(position="dodge",stat="identity")
cat("</td>")
cat("</tr></table>")
```

* TASK NO.2: Look at the spread of GDP per capita within the continents.

```{r}
# first we look at the range of gdp
gdp_spread <- gtbl %>%
  group_by(continent) %>%
  summarize(spread_gdpPercap = max(gdpPercap)-min(gdpPercap))
```

```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(gdp_spread)
cat("</td>")
cat("<td>")
ggplot(gdp_spread, aes(continent,spread_gdpPercap,fill=spread_gdpPercap))+
  geom_bar(position="dodge",stat="identity")+
  theme_bw()
cat("</td>")
cat("</tr></table>")
```

```{r}
#  Then look at the sd and iqr of the data of GDP per capita within the continents. 
gdp_spread2 <- gtbl %>%
  group_by(continent) %>%
  summarize(sd_gdp = sd(gdpPercap), iqr_gdp = IQR(gdpPercap))
gdp_spread2.r = melt(gdp_spread2)
```

```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(gdp_spread2)
cat("</td>")
cat("<td>")
ggplot(gdp_spread2.r, aes(continent,value,fill=variable))+
  geom_bar(position="dodge",stat="identity")+
  theme_bw()
cat("</td>")
cat("</tr></table>")
```

```{r}
# also box plot is a good way to see the spread of data
ggplot(gtbl, aes(continent, gdpPercap))+
  geom_boxplot(aes(fill = continent), outlier.shape = NA)+
  geom_jitter(alpha = 0.1, position = position_jitter(width = 0.1))+
  xlab("Continent")+
  ylab("GDP per capita")+
  ylim(c(0,5e4))+
  theme_bw()
```

* TASK NO.3: Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. 
```{r}
# Compute 90% trimmed mean of life expectancy for different years.
lifeExp_tmean <- gtbl %>%
  group_by(year) %>%
  summarize(tmean_lifeExp= mean(lifeExp,trim = 0.05))
```

```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(lifeExp_tmean)
cat("</td>")
cat("<td>")
ggplot(lifeExp_tmean, aes(year,tmean_lifeExp))+
  ggtitle("Trimmed average over year")+
  geom_point(color="red")+
  geom_line(color="blue")+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```

```{r}
# a weighted mean, weighting by population. 
lifeExp_wmean <- gtbl %>%
  group_by(year) %>%
  summarize(wmean_lifeExp= weighted.mean(lifeExp,pop))
```

```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(lifeExp_wmean)
cat("</td>")
cat("<td>")
ggplot(lifeExp_wmean, aes(year,wmean_lifeExp))+
  ggtitle("Weighted average over year")+
  geom_point(color="red")+
  geom_line(color="blue")+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```

* TASK NO4: How is life expectancy changing over time on different continents?

```{r}
# use weighted average lifeExp here
lifeExp_mean <- gtbl %>%
  group_by(continent, year) %>%
  summarize(wmean_lifeExp= weighted.mean(lifeExp,pop))
```

```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(lifeExp_mean)
cat("</td>")
cat("<td>")
ggplot(lifeExp_mean, aes(year, wmean_lifeExp, colour = continent)) +
  facet_wrap( ~ continent, ncol = 2) +
  theme(legend.position = "none") +
  xlab("Year") +
  ylab("Life Expectancy")+
  ggtitle("Weighted average over year")+
  geom_point()+
  geom_line()+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```

* TASK NO5: Report the absolute and/or relative abundance of countries
  with low life expectancy over time by continent: Compute some 
  measure of worldwide life expectancy – you decide – a mean or
 median or some other quantile or perhaps your current age. 
 The determine how many countries on each continent have a life 
 expectancy less than this benchmark, for each year.
 
```{r}
# use median as benchmark
benchmark<-median(gtbl$lifeExp) 
lifeExp_abu<-gtbl %>%
  group_by(continent, year) %>%
  filter(lifeExp < benchmark) %>%
  summarize(n_countries = n_distinct(country))
```
```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(lifeExp_abu)
cat("</td>")
cat("<td>")
ggplot(lifeExp_abu, aes(year,n_countries,colour = continent))+
  facet_wrap( ~ continent, ncol = 2) +
  geom_point()+
  geom_line()+
  ggtitle("NO. of countries with lifeExp under median")+
  theme_bw()
cat("</td>")
cat("</tr></table>")
```

* TASK NO6: Find countries with interesting stories.

```{r}
gtbl %>%
  filter(continent == "Asia") %>%
  select(year, country, lifeExp) %>%
  arrange(year) %>%
  group_by(year) %>%
  filter(min_rank(desc(lifeExp)) < 2 | min_rank(lifeExp) < 2)
```

We see that (min = Afghanistan, max = Japan) is the most frequent result.
```{r}
#Compare Afghanistan, Japan and world average
data_avg<- gtbl %>%
  group_by(year) %>%
  summarize(lifeExp= mean(lifeExp))
data_life<-tbl_df(data.frame(year = data_avg$year, 
                             Jap = filter(gtbl, country == "Japan")$lifeExp,
                             Afg = filter(gtbl, country == "Afghanistan")$lifeExp,
                             Avg = data_avg$lifeExp))
```
```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(data_life)
cat("</td>")
cat("<td>")
ggplot(data=data_life)+
  geom_line(aes(x = year, y=Jap,colour = "Jap"))+
  geom_line(aes(x = year, y=Afg,colour = "Afg"))+
  geom_line(aes(x = year, y=Avg,colour = "Avg"))+
  xlab("Year") + ylab("Life Expectancy")+ggtitle("Japan vs Afghanistan")+
  scale_colour_manual("Life Exp",values = c("Jap" = "red","Afg" = "blue","Avg" = "yellow"))+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```


Next, we want to find the country experiencing the sharpest 5-year drop in life expectancy.

```{r}
gtbl %>%
  group_by(continent, country) %>%
  select(country, year, continent, lifeExp) %>%
  mutate(le_delta = lifeExp - lag(lifeExp)) %>%
  summarize(worst_le_delta = min(le_delta, na.rm = TRUE)) %>%
  filter(min_rank(worst_le_delta) < 2) %>%
  arrange(worst_le_delta)
```

For above five countries, we have:
```{r}
data_drop <- tbl_df(data.frame(year = data_avg$year, 
                               Rwa = filter(gtbl, country == "Rwanda")$lifeExp,
                               Cam = filter(gtbl, country == "Cambodia")$lifeExp,
                               ES  = filter(gtbl, country == "El Salvador")$lifeExp,
                               Mon = filter(gtbl, country == "Montenegro")$lifeExp,
                               Aus = filter(gtbl, country == "Australia")$lifeExp,
                               Avg = data_avg$lifeExp))
```
```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(data_drop)
cat("</td>")
cat("<td>")
ggplot(data=data_drop)+
  geom_line(aes(x = year, y=Rwa,colour = "Rwa"))+
  geom_line(aes(x = year, y=Cam,colour = "Cam"))+
  geom_line(aes(x = year, y=ES,colour = "Es"))+
  geom_line(aes(x = year, y=Mon,colour = "Mon"))+
  geom_line(aes(x = year, y=Aus,colour = "Aus"))+
  geom_line(aes(x = year, y=Avg,colour = "Avg"))+
  xlab("Year") + ylab("Life Expectancy")+ggtitle("Great Drop")+
  scale_colour_manual("Life Exp",values = c("Rwa" = "red","Cam" = "blue",
                                            "Avg" = "yellow","Es" = "green",
                                            "Mon" = "black", "Aus" = "pink"))+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```

We also want a special analysis of Rwanda.
```{r}
# analysis for Rwanda
data_Rwa = gtbl %>% filter(country == "Rwanda")
```
```{r, results='asis', echo=FALSE, out.extra=''}
cat("<table class='container'><tr>")
cat("<td>")
kable(data_Rwa)
cat("</td>")
cat("<td>")
ggplot(data_Rwa, aes(year, pop)) +
  ggtitle("Rwanda Population")+
  geom_point(color="red")+
  geom_line(color="blue")+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
ggplot(data_Rwa, aes(year, gdpPercap)) +
  ggtitle("Rwanda gdpPercap")+
  geom_point(color="red")+
  geom_line(color="blue")+
  theme_bw()+
  theme(plot.title = element_text(lineheight=.8, face="bold"))
cat("</td>")
cat("</tr></table>")
```
We notice that population and Gdp also experienced a big decrease in 1990s. After googling, we think the reason must be [_Rwandan Genocide_](http://en.wikipedia.org/wiki/Rwandan_Genocide).

# My experience and workflow
1. `dplyr` is indeed a power tool for the analysis. Some of its grammer are similar that of `sql`. With some experience of using sql, I think I am really quick in understanding the functions in `dplyr`.
2. This time we continue our application of `ggplot`. One preblem I meet is to draw a bar graph with the data in two columns side-by-side. Although I have already learned about `position="dodge"`, it still took me some time as I finally found I need to resharp the data. See [this](http://stackoverflow.com/questions/13250872/reshaping-data-to-plot-in-r-using-ggplot2) on stackoverflow for more detail.
3. To acheive the task of put a figure and relevant table right next to each other. It need some code in html. The following code may help if anyone need
```{r}
#```{r, results='asis', echo=FALSE, out.extra=''}
#cat("<table class='container'><tr>")
#cat("<td>")
#kable(data_Rwa)
#cat("</td>")
#cat("<td>")
#ggplot(data_Rwa, aes(year, pop)) +
#  ggtitle("Rwanda Population")+
#  geom_point(color="red")+
#  geom_line(color="blue")+
#  theme_bw()+
#  theme(plot.title = element_text(lineheight=.8, face="bold"))
#ggplot(data_Rwa, aes(year, gdpPercap)) +
#  ggtitle("Rwanda gdpPercap")+
#  geom_point(color="red")+
#  geom_line(color="blue")+
#  theme_bw()+
#  theme(plot.title = element_text(lineheight=.8, face="bold"))
#cat("</td>")
#cat("</tr></table>")
#```
```